---
title: "Stat Rethinking Chapter 4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up

Load required libraries

```{r warning=FALSE, message=FALSE}
library(rethinking, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(DT, quietly = TRUE)
```

# Chapter 4

## Q4H1

The weights below were recorded from the !Kung census but heights were not recorded for these individuals.  Provide predicted heights and 89% intervals (either HPDI or PI) for each of these individuals.  That is, fill the table using model based predictions.

```{r}
kung_extra <- data.frame(i=1:5, 
                         weight=c(46.95, 43.72, 64.78, 32.59, 54.63),
                         height=rep(NA,5),
                         pi_89_lower=rep(NA,5),
                         pi_89_upper=rep(NA,5))
DT::datatable(kung_extra)                         
```

Make a model using adults:

```{r}
data(Howell1)
d_adults <- Howell1[Howell1$age >= 18,]

m4h0_flist <- alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + b*weight ,
        a ~ dnorm( 178 , 30 ) ,
        b ~ dnorm( 0 , 10 ) ,
        sigma ~ dunif( 0 , 50 )
)

m4h0 <- rethinking::map( m4h0_flist , data=d_adults )

```

Now do the predictions:

```{r}
m4h0.mu <- link( m4h0 , data=list(weight=kung_extra$weight ))
m4h0.mu.mean <- apply( m4h0.mu , 2 , mean )
m4h0.mu.HPDI <- apply( m4h0.mu , 2 , HPDI , prob=0.89 )

```
 Put the predictions in the data frame:
 
```{r}
kung_extra$height <- round(m4h0.mu.mean,1)
kung_extra$pi_89_lower <- round(m4h0.mu.HPDI[1,],1)
kung_extra$pi_89_upper <- round(m4h0.mu.HPDI[2,],1)
DT::datatable(kung_extra)
 
```

## Q4H2

Select rows in Howell dataset below 18 years of age - should get 192 rows

```{r}
data("Howell1") 
d_u18 <- Howell1[Howell1$age< 18,]
nrow(d_u18)
```

###Q4H2a

Fit a linear regression using map, and using the approximate mean for the mean of the intercept prior.

```{r}
m4h1_flist <- alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + b*weight ,
        a ~ dnorm( 100 , 30 ) ,
        b ~ dnorm( 0 , 10 ) ,
        sigma ~ dunif( 0 , 50 )
)

m4h1 <- rethinking::map( m4h1_flist , data=d_u18 )
```

View the coefficients:
```{r}
precis( m4h1 )
```

So for every 10 units increase in weight, the child gets 27.2 units taller.

###Q42Hb
Plot the raw data plus the MAP regression line and 89% HPDI for the mean.  Also superimpose the 89% HPDI for the predicted heights.
```{r}
#use link to generate MAP line and its HPDI
m4h1.weight.seq <- seq( from=2, to=50 , length.out=49 )
m4h1.mu <- link( m4h1 , data=list(weight=m4h1.weight.seq ))
m4h1.mu.mean <- apply( m4h1.mu , 2 , mean )
m4h1.mu.HPDI <- apply( m4h1.mu , 2 , HPDI , prob=0.89 )

#use sim to generate HPDI for predicted heights
m4h1.sim.height <- sim( m4h1 , data=list(weight=m4h1.weight.seq ) )
m4h1.height.HPDI <- apply( m4h1.sim.height , 2 , HPDI , prob=0.89 )

#make a data frame of model predictions for plotting
m4h1_model_plot_dat <- data.frame(weight=m4h1.weight.seq, 
                             height=m4h1.mu.mean, 
                             mu_pi_lower=m4h1.mu.HPDI[1,],
                             mu_pi_upper=m4h1.mu.HPDI[2,],
                             height_pi_lower=m4h1.height.HPDI[1,],
                             height_pi_upper=m4h1.height.HPDI[2,])

#do plot with ggplot2
library(ggplot2)
ggplot(d_u18, aes(weight, height)) + 
    geom_point() +
    geom_line(data=m4h1_model_plot_dat) +
    geom_ribbon(data=m4h1_model_plot_dat, aes(x=weight, ymin=mu_pi_lower, ymax=mu_pi_upper), 
                alpha=0.4) +
    geom_ribbon(data=m4h1_model_plot_dat, aes(x=weight, ymin=height_pi_lower, ymax=height_pi_upper), 
                alpha=0.2) +
    theme_bw()
```

### Q42Hc
Model gives straight line but relationship doesn't seem to be entirely linear - ie the predictoins are too high when weight is low and too low when weight is high.  A model including a quadratic term may be better.

## Q4H3

Fit a model including the log term:

```{r}
m4h3_flist <- alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + b*log(weight) ,
        a ~ dnorm( 178 , 100 ) ,
        b ~ dnorm( 0 , 100 ) ,
        sigma ~ dunif( 0 , 50 )
)

m4h3 <- rethinking::map( m4h3_flist , data=Howell1 )
```

Look at coefficients:

```{r}
precis(m4h3)
```
For every log unit of weight, height increases by 47 units.

### Q4H3b

```{r}
#use link to generate MAP line and its HPDI
m4h3.weight.seq <- seq( from=2, to=65 , length.out=64 )
m4h3.mu <- link( m4h3 , data=list(weight=m4h3.weight.seq ))
m4h3.mu.mean <- apply( m4h3.mu , 2 , mean )
m4h3.mu.HPDI <- apply( m4h3.mu , 2 , HPDI , prob=0.97 )

#use sim to generate HPDI for predicted heights
m4h3.sim.height <- sim( m4h3 , data=list(weight=m4h3.weight.seq ) )
m4h3.height.HPDI <- apply( m4h3.sim.height , 2 , HPDI , prob=0.97 )

#make a data frame of model predictions for plotting
m4h3_model_plot_dat <- data.frame(weight=m4h3.weight.seq, 
                             height=m4h3.mu.mean, 
                             mu_pi_lower=m4h3.mu.HPDI[1,],
                             mu_pi_upper=m4h3.mu.HPDI[2,],
                             height_pi_lower=m4h3.height.HPDI[1,],
                             height_pi_upper=m4h3.height.HPDI[2,])

#do plot with ggplot2
library(ggplot2)
ggplot(Howell1, aes(weight, height)) + 
    geom_point() +
    geom_line(data=m4h3_model_plot_dat) +
    geom_ribbon(data=m4h3_model_plot_dat, aes(x=weight, ymin=mu_pi_lower, ymax=mu_pi_upper), 
                alpha=0.4) +
    geom_ribbon(data=m4h3_model_plot_dat, aes(x=weight, ymin=height_pi_lower, ymax=height_pi_upper), 
                alpha=0.2) +
    theme_bw()

```




