
```{r}
library(ggplot2)
```

## Easy Questions

### Q.2E1
_the probability of rain on Monday_ corresponds to Pr(rain|Monday) - ie the probability of rain given that it is a monday

### Q.2E2
Pr(Monday|rain) corresponds to the probability that it is a Monday given that it is raining.

### Q.2E3
_the probability that it is Monday, given that it is raining_ corresponds to:
Pr(Monday|rain)
Pr(rain|Monday) Pr(Monday)/ Pr(rain)

### Q.2E4
_that probability of water_ is the probability that what the ball is thrown and caught, that the index finger of the catcher lies on water.  This is equivalent to the proportion of water on the globe.

## Medium Quesions

### Q.2M1
Compute and plot the grid approximate posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for p. WWW, WWWL, LWWLWWW

```{r}
#define grid
p_grid <- seq(from=0, to=1, length.out=1000)

#define prior
prior <- rep(1,1000)

#compute posterior for WWW
lik1 <- dbinom(3, size=3, prob=p_grid) #compute likelihood at each point in grid - ie 3/3 water
posterior_unstd1 <- lik1 * prior #compute product of likelihood and prior
posterior1 <- posterior_unstd1/sum(posterior_unstd1)  #standardise the posterior so it sums to 1
qplot(p_grid, posterior1) + geom_point() + ggtitle('WWW') + theme_bw()

#compute posterior for WWWL
lik2 <- dbinom(3, size=4, prob=p_grid) #compute likelihood at each point in grid - ie 3/4 water
posterior_unstd2 <- lik2 * prior #compute product of likelihood and prior
posterior2 <- posterior_unstd2/sum(posterior_unstd2)  #standardise the posterior so it sums to 1
qplot(p_grid, posterior2) + geom_point() + ggtitle('WWWL') + theme_bw()

#compute posterior for LWWLWWW
lik3 <- dbinom(5, size=7, prob=p_grid) #compute likelihood at each point in grid - ie 5/7 water
posterior_unstd3 <- lik3 * prior #compute product of likelihood and prior
posterior3 <- posterior_unstd3/sum(posterior_unstd3)  #standardise the posterior so it sums to 1
qplot(p_grid, posterior3) + geom_point() + ggtitle('LWWLWWW') + theme_bw()

```

### Q.2M2
Compute and plot the grid approximate posterior distributions for the above when the prior is zero when p<0.5 and constant when p>0.5

```{r}
#define grid
p_grid <- seq(from=0, to=1, length.out=1000)

#define prior
prior <- as.numeric(p_grid > 0.5)

#compute posterior for WWW
lik1 <- dbinom(3, size=3, prob=p_grid) #compute likelihood at each point in grid - ie 3/3 water
posterior_unstd1 <- lik1 * prior #compute product of likelihood and prior
posterior1 <- posterior_unstd1/sum(posterior_unstd1)  #standardise the posterior so it sums to 1
qplot(p_grid, posterior1, geom = 'line') + ggtitle('WWW') + theme_bw()

#compute posterior for WWWL
lik2 <- dbinom(3, size=4, prob=p_grid) #compute likelihood at each point in grid - ie 3/4 water
posterior_unstd2 <- lik2 * prior #compute product of likelihood and prior
posterior2 <- posterior_unstd2/sum(posterior_unstd2)  #standardise the posterior so it sums to 1
qplot(p_grid, posterior2, geom = 'line') + ggtitle('WWWL') + theme_bw()

#compute posterior for LWWLWWW
lik3 <- dbinom(5, size=7, prob=p_grid) #compute likelihood at each point in grid - ie 5/7 water
posterior_unstd3 <- lik3 * prior #compute product of likelihood and prior
posterior3 <- posterior_unstd3/sum(posterior_unstd3)  #standardise the posterior so it sums to 1
qplot(p_grid, posterior3, geom = 'line') + ggtitle('LWWLWWW') + theme_bw()


```

### Q.2M3
Suppose there are two globes, one for Earth and one for Mars. The Earth globe is 70% covered in water. The Mars globe is 100% land. Further suppose that one of these globes—you don't know which—was tossed in the air and produced a “land” observation. Assume that each globe was equally likely to be tossed. Show that the posterior probability that the globe was the Earth, conditional on seeing “land” (Pr(Earth|land)), is 0.23.

Pr(Earth|land) = Pr(land|Earth)*Pr(Earth)/Pr(land)

Pr(land|Earth) = 0.3
Pr(land|Mars) = 1
So Pr(land) = 0.3 + 1 / 2 = 0.65
Pr(Earth) = 0.5

Thus Pr(Earth|land): 
```{r}
0.3*0.5/0.65
```

### Q.2M4
Three cards B/B, B/W, W/W.  Put in bag and shuffled.  Card pulled and B up but unknown down - probability of other side being black is 2/3.  Show.

How many ways could each card produce the data?
B/B - 2 ways
B/W - 1 way
W/W - 0 ways

'Other side being black' is equivalent to card being B/B, so probability is 2/3.

### Q.2M5
Now there are 4 cards: B/B, B/W, W/W, B/B
B/B-1 - 2 ways
B/B-2 - 2 ways
B/W - 1 way
W/W - 0 ways

Probability now 4/5

### Q 2.M6
Now for every way to pull the B/B card from the bag, there are 2 ways to pull the B/W card and 3 ways to pull the W/W card.  Show prob now 0.5.

B/B - 2 ways * 1 skew factor = 2
B/W - 1 way * 2 skew factor = 2
W/W - 0 ways * 3 skew factor = 0

Probability now 2/2 = 0.5

### Q 2.M7
Now draw 2 cards.  Second card is white face up.  Prob that first card is B/B now 0.75

B/B then B/W - 2 ways
B/B then W/W - 4 ways
B/W then B/B - 0 ways
B/W then W/W - 2 ways
W/W then B/B - 0 ways
W/W then B/W - 0 ways

Now there are 6/8 ways that first card is B/B so prob 0.75.

### Q 2.H1
Pr(twins|A) = 0.1
Pr(twins|B) = 0.2
Pr(A) = 0.5
Pr(B) = 0.5


Panda given birth to twins, prob next birth twins.  
With no prior knowledge our estimate (Pr(twins)) would be 0.15 since A/B equally likely, or more formally it's 0.1x0.5 + 0.2x0.5
But panda more likely to be species B, so probability Pr(twins) should be between 0.15 and 0.2.

First work out Pr(A|twins) and Pr(B|twins)
Pr(A|twins) = Pr(twins|A)*Pr(A)/Pr(twins):
```{r}
P_A_given_twins <- 0.1 * 0.5 / 0.15
P_A_given_twins
P_B_given_twins <- 1-P_A_given_twins
P_B_given_twins
```
We can now use these updated priors to work out a new value for Pr(twins):
Pr(twins) = Pr(twins|A) x Pr(updated A) + Pr(twins|B) x Pr(updated B)

```{r}
0.1 * P_A_given_twins + 0.2 * P_B_given_twins
```

### Q 2.H2
From 2H1 Pr(A|twins) is 0.333

### Q 2.H3
Now panda has a second birth and it is not twins.  Compute posterior probability that panda is species A

Use Bayes rule again ie:
Pr(twins|A) = 0.1
Pr(twins|B) = 0.2
Pr(A) = 0.333 (updated previously)
Pr(B) = 0.666

Pr(A|not twins) = Pr(not twins|A)*Pr(A)/Pr(not twins)
```{r}
(1-0.1) * P_A_given_twins / (1-0.1666667)
```





