---
title: "Stat Rethinking Chapter 5"
output: html_document
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up

Load required libraries

```{r warning=FALSE, message=FALSE}
library(rethinking, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(DT, quietly = TRUE)
```

# Chapter 5

## Q5H1

Load the foxes data and view:

```{r}
data(foxes)
knitr::kable(foxes[1:10,])
```

### Weight vs Area
Fit a bivariate Gaussian regression for body weight as a linear function of territory size:

```{r}
m5h1_flist <- alist(
        weight ~ dnorm( mu , sigma ) ,
        mu <- a + b*area ,
        a ~ dnorm( 4.5 , 10 ) ,
        b ~ dnorm( 0 , 100 ) ,
        sigma ~ dunif( 0 , 50 )
)

m5h1 <- rethinking::map( m5h1_flist , data=foxes )
```

Quickly look at the parameter estimates:

```{r}
precis(m5h1)
plot(precis(m5h1))
```

Now look at MAP regression line and 95% interval of the mean

```{r}
#use link to generate MAP line and its HPDI
m5h1.area.seq <- seq( from=0.5, to=7 , by = 0.1 )
m5h1.mu <- link( m5h1 , data=list(area=m5h1.area.seq ))
m5h1.mu.mean <- apply( m5h1.mu , 2 , mean )
m5h1.mu.HPDI <- apply( m5h1.mu , 2 , HPDI , prob=0.97 )

#make a data frame of model predictions for plotting
m5h1_model_plot_dat <- data.frame(area=m5h1.area.seq, 
                             weight=m5h1.mu.mean, 
                             mu_pi_lower=m5h1.mu.HPDI[1,],
                             mu_pi_upper=m5h1.mu.HPDI[2,])

#do plot with ggplot2
library(ggplot2)
ggplot(foxes, aes(area, weight)) + 
    geom_point() +
    geom_line(data=m5h1_model_plot_dat) +
    geom_ribbon(data=m5h1_model_plot_dat, aes(x=area, ymin=mu_pi_lower, ymax=mu_pi_upper), 
                alpha=0.4)  +
    theme_bw()

```

### Weight vs Area
Fit a bivariate Gaussian regression for body weight as a linear function of group size:

```{r}
m5h1b_flist <- alist(
        weight ~ dnorm( mu , sigma ) ,
        mu <- a + b*groupsize ,
        a ~ dnorm( 4.5 , 10 ) ,
        b ~ dnorm( 0 , 100 ) ,
        sigma ~ dunif( 0 , 50 )
)

m5h1b <- rethinking::map( m5h1b_flist , data=foxes )
```

Quickly look at the parameter estimates:

```{r}
precis(m5h1b)
plot(precis(m5h1b))
```

Now look at MAP regression line and 95% interval of the mean

```{r}
#use link to generate MAP line and its HPDI
m5h1b.groupsize.seq <- seq( from=1, to=9 , by = 0.2 )
m5h1b.mu <- link( m5h1b , data=list(groupsize=m5h1b.groupsize.seq ))
m5h1b.mu.mean <- apply( m5h1b.mu , 2 , mean )
m5h1b.mu.HPDI <- apply( m5h1b.mu , 2 , HPDI , prob=0.97 )

#make a data frame of model predictions for plotting
m5h1b_model_plot_dat <- data.frame(groupsize=m5h1b.groupsize.seq, 
                             weight=m5h1b.mu.mean, 
                             mu_pi_lower=m5h1b.mu.HPDI[1,],
                             mu_pi_upper=m5h1b.mu.HPDI[2,])

#do plot with ggplot2
library(ggplot2)
ggplot(foxes, aes(groupsize, weight)) + 
    geom_point() +
    geom_line(data=m5h1b_model_plot_dat) +
    geom_ribbon(data=m5h1b_model_plot_dat, aes(x=groupsize, ymin=mu_pi_lower, ymax=mu_pi_upper), 
                alpha=0.4)  +
    theme_bw()

```

### Conclusion

Neither group size nor area appears to be important for predicting fox body weight based on the bivariate regressions.

## Q5H2

In this question we want to see the effect of including both groupsize AND area in the regression.

### Fit a multiple linear regression

```{r}
m5h2_flist <- alist(
        weight ~ dnorm( mu , sigma ) ,
        mu <- a + bG*groupsize + bA*area ,
        a ~ dnorm( 4.5 , 10 ) ,
        bG ~ dnorm( 0 , 100 ) ,
        bA ~ dnorm( 0 , 100 ) ,
        sigma ~ dunif( 0 , 50 )
)

m5h2 <- rethinking::map( m5h2_flist , data=foxes )
```

Quickly look at the parameter estimates:

```{r}
precis(m5h2)
plot(precis(m5h2))
```

Now the variables do look more important since - larger group size seems to negatively impact weight whereas larger area size positively impacts weight.

### Counterfactual plots

First explore groupsize:

```{r}
#set up prediction data
pred_data_m5h2a <- data.frame(area=mean(foxes$area),
                        groupsize=seq( from=1, to=9 , by = 0.2 ))

#predict mu from model
mu_m5h2a <- link(m5h2, data=pred_data_m5h2a, n=1e4)
mu_mean_m5h2a <- apply(mu_m5h2a, 2, mean)
mu_PI_m5h2a <- apply(mu_m5h2a, 2, PI, prob=0.975 )

#make a data frame of model predictions for plotting
m5h2a_model_plot_dat <- data.frame(area = pred_data_m5h2a$area,
                             groupsize=pred_data_m5h2a$groupsize, 
                             weight=mu_mean_m5h2a, 
                             mu_pi_lower=mu_PI_m5h2a[1,],
                             mu_pi_upper=mu_PI_m5h2a[2,])

#do plot with ggplot2
ggplot(m5h2a_model_plot_dat, aes(groupsize, weight)) + 
    geom_line(data=m5h2a_model_plot_dat) +
    geom_ribbon(data=m5h2a_model_plot_dat, aes(x=groupsize, ymin=mu_pi_lower, ymax=mu_pi_upper), 
                alpha=0.4)  +
    theme_bw()


```

Now area:

First explore groupsize:

```{r}
#set up prediction data
pred_data_m5h2b <- data.frame(area=seq( from=0.5, to=7 , by = 0.1 ),
                        groupsize=mean(foxes$groupsize))

#predict mu from model
mu_m5h2b <- link(m5h2, data=pred_data_m5h2b, n=1e4)
mu_mean_m5h2b <- apply(mu_m5h2b, 2, mean)
mu_PI_m5h2b <- apply(mu_m5h2b, 2, PI, prob=0.975 )

#make a data frame of model predictions for plotting
m5h2b_model_plot_dat <- data.frame(area = pred_data_m5h2b$area,
                             area=pred_data_m5h2b$groupsize, 
                             weight=mu_mean_m5h2b, 
                             mu_pi_lower=mu_PI_m5h2b[1,],
                             mu_pi_upper=mu_PI_m5h2b[2,])

#do plot with ggplot2
ggplot(m5h2b_model_plot_dat, aes(area, weight)) + 
    geom_line(data=m5h2b_model_plot_dat) +
    geom_ribbon(data=m5h2b_model_plot_dat, aes(x=area, ymin=mu_pi_lower, ymax=mu_pi_upper), 
                alpha=0.4)  +
    theme_bw()


```

### Conclusion

Including both variables in the model shows that area has a positive impact on weight whilst groupsize has a negative impact on weight.  THe size of the effects is bigger than when the variables were examined in seperate models.  This is due to the masking effect where two variables with effects in opposite directions can mask eachother out unless considered together. 


